Multifluid model of one-dimensional force-free current sheet equilibrium

### 2024-10-04 article.qmd
update: SPARTHRB presentation

---
title: On a Alfvenicity of solar wind discontinuities
authors:
  - name: Zijin Zhang
    email: zijin@ucla.edu
    orcid: 0000-0002-9968-067X
    affiliation: University of California, Los Angeles
  - name: Sergei Kamaletdinov
    email: sergei2033@ucla.edu
    orcid: 0000-0002-2261-0331
    affiliation: University of California, Los Angeles
  - name: Xiaofei Shi
    email: sxf1698@ucla.edu
    orcid: 0000-0003-3367-5074
    affiliation: University of California, Los Angeles
  - name: Anton V. Artemyev
    email: aartemyev@igpp.ucla.edu
    orcid: 0000-0001-8823-4474
    affiliation: University of California, Los Angeles
  - name: Marco Velli
    email: mvelli@ucla.edu
    orcid: 0000-0002-2381-3106
    affiliation: University of California, Los Angeles
  - name: Vassilis Angelopoulos
    orcid: 0000-0001-7024-1561
    affiliation: University of California, Los Angeles
code-links:
  - text: Overleaf
    href: https://www.overleaf.com/project/668273f5d14067a22517615c
  - text: Roam Research
    href: https://roamresearch.com/#/offline/Research/page/5EiGYXsXE
  - text: Zotero
    href: zotero://select/library/collections/DZHNUFNV
abstract: "Solar wind plasma flow is characterized by intense, large-amplitude current sheets, also known as solar wind discontinuities. These regions are believed to play a crucial role in magnetic reconnection, solar wind ion heating, and energetic particle scattering. Although solar wind discontinuities share several properties with the classical MHD model of rotational discontinuities, they exhibit more complex characteristics that include many kinetic properties, going beyond this model. One of the most puzzling properties of these discontinuities is their Alvenicity, which describes the relationship between Alfvén and plasma velocity jumps across discontinuities. Although the high correlation of these jumps suggests the rotational nature of discontinuities, the significant differences in jump amplitudes indicate the necessity of kinetic mechanisms to describe the stress balance within discontinuities. Spacecraft observations show that plasma thermal anisotropy alone may not sufficiently explain this difference and achieve stress balance. In this study, we present a theoretical model in which this difference is explained by the kinetics of interpenetrating ion beams crossing the discontinuity. Our model successfully describes the observed properties of discontinuities within the inner heliosphere."
---

:::{.hide}
\newcommand{\vect}[1]{\mathbf{#1}}
\DeclareMathOperator{\sech}{sech}
:::

@@

1. Observation features of solar wind discontinuities
2. The necessity of kinetic models to describe solar wind discontinuities
3. Current theoretical models of solar wind discontinuities and their limitations
  - Harris current sheet @harrisPlasmaSheathSeparating1962

Observations of magnetic discontinuities exhibit Alfvenic properties, with the plasma velocity variation correlating well with magnetic field components variations. However, the amplitude of the plasma velocity jump is often significantly smaller than the Alfvén velocity jump, indicating a more complicated picture than the classical MHD model of rotational discontinuities. @fig-examples show observations of  magnetic discontinuities in different regions of the heliosphere, PSP within 0.2 AU, Wind upstream of the Earth's bow shock, and ARTEMIS in the Earth's magnetotail.  Previously, this discrepancy has been interpreted as a possible contribution of anisotropy that decreases the Alfvén velocity jump, or a discontinuity non-stationarity due to residual magnetic energy, or an ion nonadiabatic interaction with intense (thin) discontinuities that shapes the nongyrotropic ion distribution to include a nondiagonal term of the pressure tensor, with a cross-discontinuity gradient decreasing the Alfvén velocity jump @artemyevIonNongyrotropySolar2020.

These differences could not be simply explained by the aniostropy of the plasma thermal pressure alone (assuming a Maxwellian distribution that fits both the parallel and perpendicular temperatures) and suggest that kinetic mechanisms are necessary to describe the stress balance within discontinuities.

@@

Three examples of solar wind discontinuities observed by Parker Solar Probe (PSP), ARTEMIS and Wind spacecraft with sub-Alfvenic flow velocity.
:::

Force-free @harrisonRemarksOnedimensionalForcefree2009 (A necessary condition on the pseudopotential plasma pressure to allow for force-free 1D VM equilibria is formulated.)

@@

@@

$$
d B_z / d z = J_z = \sum_α n_α u_{α z} - n_e u_{e z} = n u_z - Γ_e
$$ {#eq-Jz}

@@

$$
- B_0 \sin θ θ' = \sum n_α u_{α} \sin θ_α - Γ_e B_0 \sin θ / B_z
% Λ_y n_1 u_1 \sin θ_1 - Λ_z Γ_1 B_0 \sin θ / B_z
$$

@@

For the simplest situation where we have two group of ions of the same density but opposite bulk velocity, i.e., $n_1 = n_2, u_1 = -u_2$, the system could be determined given $κ$, $L$, and $B_0$. The system profile of magnetic field, plasma density, and plasma velocity for one specific case ($δ_1=1, c_1=1/\sqrt{2}, B_0 = 2$) is plotted below @fig-profiles. The specific profiles are chosen to have zero electron current across the current sheet and zero $B_y$ in the asymptotic limit, corresponding to a 180° rotation of the magnetic field across the current sheet. However, in general, we would expect the electron current to be non-zero.

@@

(a). Magnetic field, ion density, and ion bulk velocity for $n_1 = n_2, u_1 = -u_2$ and $δ_1=1, c_1=1, B_0 = 2$. (b). Current density profiles for the same case.
:::

@@

The asymptotic velocity for the each species, bulk velocity and normalized one by Alfvén velocity are plotted below @fig-vRatios and @fig-vRatiosEx2 for the case of $\lambda_{z,2} = - 1$ (two groups of opposite velocity in z direction) and $\lambda_{z,2} = - 1/2$ respectively. For the case of $\lambda_{z,2} = - 1$ as shown in @fig-vRatiosEx1, the normalized velocity reaches zero when $n_1(∞) = 0.5$. The bulk velocity of each species is the same in the asymptotic limit and does not depend on $n_1(∞)$. However, this does not generally hold when $\lambda_{z,2} \neq - 1$ as shown in @fig-vRatiosEx2 for the case of $\lambda_{z,2} = - 1/2$.. The normalized velocity reaches zero when $n_1(∞) = 1/3$, and the bulk velocity of each species differs in the asymptotic limit and depends on $n_1(∞)$.

@@

Asymptotic velocity for each species, bulk velocity and normalized one by Alfvén velocity for the case of $\lambda_{z,2} = - 1$ (a) and $\lambda_{z,2} = - 1/2$ (b).
:::
<!-- ::: -->

@@

In non-dimensionalized form
$$
\begin{aligned}
B_x(z) &= B_0 \sin \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
B_y(z) &= B_0 \cos \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
n(z) &= n(\infty ) + \frac{\kappa}{V_a} \frac{∑_α Γ_α}{1+(z/L)^2}
\\
J_x(z) &= \frac{\kappa B_0}{\mu_0 d_i}\frac{1}{1+(z/L)^2} \sin \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
J_y(z) &= \frac{\kappa B_0}{\mu_0 d_i}\frac{1}{1+(z/L)^2} \cos \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\end{aligned}
$$

>>>>

---
title: On the Alfvenicity of solar wind discontinuities
authors:
  - name: Zijin Zhang
    email: zijin@ucla.edu
    orcid: 0000-0002-9968-067X
    affiliation: University of California, Los Angeles
  - name: Sergei Kamaletdinov
    email: sergei2033@ucla.edu
    orcid: 0000-0002-2261-0331
    affiliation: University of California, Los Angeles
  - name: Xiaofei Shi
    email: sxf1698@ucla.edu
    orcid: 0000-0003-3367-5074
    affiliation: University of California, Los Angeles
  - name: Anton V. Artemyev
    email: aartemyev@igpp.ucla.edu
    orcid: 0000-0001-8823-4474
    affiliation: University of California, Los Angeles
  - name: Marco Velli
    email: mvelli@ucla.edu
    orcid: 0000-0002-2381-3106
    affiliation: University of California, Los Angeles
  - name: Vassilis Angelopoulos
    orcid: 0000-0001-7024-1561
    affiliation: University of California, Los Angeles
code-links:
  - text: Overleaf
    href: https://www.overleaf.com/project/668273f5d14067a22517615c
  - text: Roam Research
    href: https://roamresearch.com/#/offline/Research/page/5EiGYXsXE
  - text: Zotero
    href: zotero://select/library/collections/DZHNUFNV
abstract: "Solar wind plasma flow is characterized by intense, large-amplitude current sheets, also known as solar wind discontinuities. These regions are believed to play a crucial role in magnetic reconnection, solar wind ion heating, and energetic particle scattering. Although solar wind discontinuities share several properties with the classical MHD model of rotational discontinuities, they exhibit more complex characteristics that include many kinetic properties, going beyond this model. One of the most puzzling properties of these discontinuities is their Alvenicity, which describes the relationship between Alfvén and plasma velocity jumps across discontinuities. Although the high correlation of these jumps suggests the rotational nature of discontinuities, the significant differences in jump amplitudes indicate the necessity of kinetic mechanisms to describe the stress balance within discontinuities. Spacecraft observations show that plasma thermal anisotropy alone may not sufficiently explain this difference and achieve stress balance. In this study, we present a theoretical model in which this difference is explained by the kinetics of interpenetrating ion beams crossing the discontinuity. Our model successfully describes the observed properties of discontinuities within the inner heliosphere."
---

:::{.hide}
\newcommand{\vect}[1]{\mathbf{#1}}
:::

@@

- Observation features of solar wind discontinuities
- Importance of solar wind discontinuities
- The necessity of kinetic models to describe solar wind discontinuities
- Current theoretical models of solar wind discontinuities and their limitations
  - Harris current sheet @harrisPlasmaSheathSeparating1962

<!-- Observations of magnetic discontinuities exhibit Alfvenic properties, with the plasma velocity variation correlating well with magnetic field components variations. However, the amplitude of the plasma velocity jump is often significantly smaller than the Alfvén velocity jump, indicating a more complicated picture than the classical MHD model of rotational discontinuities. @fig-examples show observations of  magnetic discontinuities in different regions of the heliosphere, PSP within 0.2 AU, Wind upstream of the Earth's bow shock, and ARTEMIS in the Earth's magnetotail.  Previously, this discrepancy has been interpreted as a possible contribution of anisotropy that decreases the Alfvén velocity jump, or a discontinuity non-stationarity due to residual magnetic energy, or an ion nonadiabatic interaction with intense (thin) discontinuities that shapes the nongyrotropic ion distribution to include a nondiagonal term of the pressure tensor, with a cross-discontinuity gradient decreasing the Alfvén velocity jump @artemyevIonNongyrotropySolar2020.

These differences could not be simply explained by the aniostropy of the plasma thermal pressure alone [@shenComparingPlasmaAnisotropy2024] (assuming a Maxwellian distribution that fits both the parallel and perpendicular temperatures) and suggest that kinetic mechanisms are necessary to describe the stress balance within discontinuities. -->

@@

Three examples of current sheets observed by Parker Solar Probe (PSP), ARTEMIS and Wind spacecraft with sub-Alfvenic flow velocity.
:::

@@

$$
\mathbf {\nabla } \times \mathbf {B} =\mu _{0}\mathbf {J}
$$ {#eq-Ampere}

$$
∇ · \mathbf{B} = 0
$$

@@

$$
0 = J_z = \sum_α n_α u_{α z} - n_e u_{e z} = n u_z - Γ_e
$$ {#eq-Jz}

@@

$$
- B_0 \sin θ θ' = \sum n_α u_{α} \sin θ_α - Γ_e B_0 \sin θ / B_z
$$

@@

For the simplest situation where we have two group of ions of the same density but opposite bulk velocity, i.e., $n_1 = n_2, u_1 = -u_2$, the system could be determined given $κ$, $L$, and $B_0$. The system profile of magnetic field, plasma density, and plasma velocity for one specific case ($L=1, κ=1, B_0 = 2$) is plotted below @fig-profiles. The specific profiles are chosen to have zero electron current across the current sheet and zero $B_y$ in the asymptotic limit, corresponding to a 180° rotation of the magnetic field across the current sheet. However, in general, we would expect the electron current to be non-zero.

@@

(a). Magnetic field, ion density, and ion bulk velocity for $n_1 = n_2, u_1 = -u_2$ and $L=1, κ=1, B_0 = 2$. (b). Current density profiles for the same case.
:::

@@

The asymptotic velocity for the each species, bulk velocity and normalized one by Alfvén velocity are plotted below @fig-vRatios and @fig-vRatiosEx2 for the case of $u_{1,z} = - u_{2,z}$ (two groups of opposite velocity in z direction) and $u_{1,z} = - 2 u_{2,z}$ respectively. For the case of $u_{1,z} = - u_{2,z}$ as shown in @fig-vRatiosEx1, the normalized velocity reaches zero when $n_1(∞) = 0.5$. The bulk velocity of each species is the same in the asymptotic limit and does not depend on $n_1(∞)$. However, this does not generally hold when $u_{1,z} \neq - u_{2,z}$ as shown in @fig-vRatiosEx2 for the case of $u_{1,z} = - 2 u_{2,z}$. The normalized velocity reaches zero when $n_1(∞) = 1/3$, and the bulk velocity of each species differs in the asymptotic limit and depends on $n_1(∞)$.

@@

Asymptotic velocity for each species, bulk velocity and normalized one by Alfvén velocity for the case of $u_{1,z} = - u_{2,z}$ (a) and $u_{1,z} = - 2 u_{2,z}$ (b).
:::
<!-- ::: -->

@@

In non-dimensionalized form
$$
\begin{aligned}
B_x(z) &= B_0 \sin \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
B_y(z) &= B_0 \cos \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
n(z) &= n_{\infty } + \frac{\kappa}{V_A} \frac{∑_α Γ_α}{1+(z/L)^2}
\\
J_x(z) &= \frac{\kappa B_0}{\mu_0 d_i}\frac{1}{1+(z/L)^2} \sin \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
J_y(z) &= \frac{\kappa B_0}{\mu_0 d_i}\frac{1}{1+(z/L)^2} \cos \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\end{aligned}
$$

### 2024-10-04 index.qmd
update manuscripts and notebooks
update: SPARTHRB presentation

---
title: "Multifluid model of a one-dimensional steady state force-free current sheet"
format:
  html: default
  pptx:
    slide-level: 3
    incremental: true
code-links:
  - text: Tana
    href: https://app.tana.inc?nodeid=OzzX0sVA4_6X
  - text: Zotero
    href: zotero://select/library/collections/DZHNUFNV
---

:::{.hide}
\newcommand{\vect}[1]{\mathbf{#1}}
:::

## Introduction

### Motivation

In MHD theory, for rotational rotational discontinuity, we have this following relationship

$$
[[\vec{U}]]=\left(\frac{\xi \rho}{\mu_0}\right)^{1 / 2}\left[\left[\frac{\vec{B}}{\rho}\right]\right]
$$

where $\xi \equiv 1-\frac{p_{\parallel}-p_{\perp}}{B^2 / \mu_0}$.

@@

However, observations of solar wind discontinuities reveals discrepancies between

- Alfven velocity and plasma velocity change across discontinuities
- Anisotropic MHD theory-predicted and directly measured ion anisotropies

Multifluid theory is proposed to address these discrepancies. The basic idea is that we could have zero bulk velocity with non zero pressure.

#### MHD in a nutshell

$$
\begin{aligned}
& \frac{\partial \rho}{\partial t}+(\vec{V} \cdot \nabla) \rho+\rho \nabla \cdot \vec{V}=0 \\
& \rho\left[\frac{\partial \vec{V}}{\partial t}+(\vec{V} \cdot \nabla) \vec{V}\right]+\nabla p=\vec{J} \times \vec{B} \\
& p=\alpha \rho^\gamma \\
& \nabla \times \vec{B}=\mu_0 \vec{J} \\
& \frac{\partial \vec{B}}{\partial t}=\nabla \times(\vec{V} \times \vec{B})
\end{aligned}
$$

### Previous Work

On the extreme of multiple fluid theory, we have the kinetic Harris model for the current sheet.

- no normal field, $B_z = 0$
- uniform cross-tail drift velocity, $u_y$
- vanishing particle density far from the sheet

@@

For the symmetric three-fluid system with $B_y=0$,

@@

Here, we are interested in the case of a force-free current sheet and start from the simplest case of two ion species.

## Plasma-Field Equations for a One-Dimensional Multifluid

Consider a collisionless, steady state plasma composed of multiple fluid groups, and follow the notation in @steinhauerMultifluidModelOnedimensional2008, we have $3N$ equations for $N$ ion species:

@@

$$
d B_y / d z = - J_x = - \sum_α n_α u_{α x} + n_e u_{e x} = -n u_x+\Gamma_e B_x / B_z
$${#eq-Jx}

$$
d B_x / d z = J_y = \sum_α n_α u_{α y} - n_e u_{e y} = n u_y-\Gamma_e B_y / B_z
$${#eq-Jy}

@@

::: {#def-1}
Express the quantities of the second ion species relative to the first species

@@

- $Γ_e = (1 + λ_z λ_n) Λ_1 =  Λ_z  Γ_1$. 
- $Λ_z$ is constant while $Λ_x, Λ_y$ are not.

:::

@@

$C_1=0$ immediately implies $θ = θ_1 + k \pi$ and $Λ_x=Λ_y$. #TODO: Prove $C_1=0$ is like a boundary condition not a equation.

@@

$$
\Rightarrow 
θ_α' = - \frac{n_α B_z}{Γ_α} \pm \frac{B_0}{u_α}
$${#eq-theta1}

- Note that $θ_α, n_α$ are dependent variables, and $u_α, B_0, B_z, Γ_α$ are constants determined by the system.
- **So given the profile of $n_α$, we could solve the above equation to get the profile of $θ_α$, thus the profile of $u_α$.**
- The derivative of $θ$ goes to zero at infinity, gives us a relation between $u_1$ and $B_0$:

  $$
  u_1 = \pm \frac{B_0 Γ_1}{B_z n_1(\infty)}
  $${#eq-u1}

@@

$$
- B_0 \sin θ θ' = Λ_y n_1 u_1 \sin θ_1 - Λ_z Γ_1 B_0 \sin θ / B_z
$$

$$
\Rightarrow 
θ' = \mp \frac{Λ_y n_1 u_1 }{B_0} + \frac{Λ_z Γ_1}{B_z}
$${#eq-theta}

By equating the above two equations, we could get a relation between $n_1$ and $Λ_y$:

$$
Λ_y = Λ_y(n_1) = -\frac{B_0 \left(-n_1 u_1 B_z^2+B_0 \Gamma _1 B_z-\Gamma _1^2 u_1 \Lambda _z\right)}{\Gamma _1 n_1 u_1^2 B_z}
$$

### Solutions

Normalize the density by $n_1(z=\infty) = 1$ and the magnetic field by $B_z = 1$.

@@

Assuming density profile $n_1(z) \to \frac{c}{\left(\frac{z}{\delta }\right)^2+1}+1$

We have

$$
\begin{aligned}
\varphi (z) &\to \frac{\pi  \Gamma _1-2 c \delta  \tan ^{-1}\left(\frac{z}{\delta }\right)}{2 \Gamma _1}
\\
B_x &\to B_0 \cos \left(\frac{\pi  \Gamma _1-2 c \delta  \tan ^{-1}\left(\frac{z}{\delta }\right)}{2 \Gamma _1}\right)
\end{aligned}
$$

@@

![](figures/J_profiles.jpg)

@@

```mathematica
Manipulate[
 Block[{B0 = B0i, c = ci},
  p1 = Plot[
    Evaluate[{n[z], Bx[z], By[z], u1x[z], u1y[z]} /. sols //. rulesn],
    {z, -zmax, zmax},
    PlotLegends -> {n, Subscript[B, x], Subscript[B, y], Subscript[u, x], Subscript[u, y]}
  ];
  Export["figures/profiles.jpg", p1];
  p1
  ],

 {{ci, 1/Sqrt[2]}, 0, 2},
 {{B0i, 2}, 0.5, 2},
 {{zmax, 10}, 5, 30}
 ]

Manipulate[
 Block[{B0 = B0i, c = ci, Λz = Λz0},
  p = Plot[
    Evaluate[{Λy - 1, Jxi, Jx, Jyi, Jy} /. sols //. 
      rulesn],
    {z, -zmax, zmax},
    PlotLegends -> {Subscript[Λ, y] - 1, Subscript[J, 
      x, i], Subscript[J, x], Subscript[J, y, i], Subscript[J, y]}, 
    PlotRange -> All];
  Export["figures/J_profiles.jpg", p];
  p
  ],

 {{ci, 1/Sqrt[2]}, 0, 2},
 {{B0i, 2}, 0.5, 2},
 {Λz0, 0, 1},
 {{zmax, 10}, 5, 30}
]
```

@@

::: {.column-screen-inset}
::: {#fig-profiles layout-nrow="1" }

![Magnetic field, ion density, and out ion bulk velocity for $λ_{n,∞}=1$ and $λ_{z,∞}=-1$.](figures/profiles_ratio=0.5.svg){#fig-profiles_ratio}

![Normalized Plasma Velocity $U_x$ Profiles for Different $λ_{n,∞}$.](figures/UxNorm.svg){#fig-uxnorm}

![Normalized Plasma Velocity $U_y$ Profiles for Different $λ_{n,∞}$.](figures/UyNorm.svg){#fig-uynorm}

:::
:::

![](figures/vRatios.svg)

## Different normalization

Normalized the quantities of the ion species

$$
n_α := λ_{αn} n_e
\\
u_{αx} := λ_{αu} B_z / \sqrt{n_{e,\infty}}
\\
u_{αy} := λ_{αu} B_z / \sqrt{n_{e,\infty}}
\\
u_{αz} := λ_{αu} B_z / \sqrt{n_{e,\infty}}
$$

$$
θ' = \mp \frac{Λ_u}{B_0} + Γ_e
$$

```mathematica
conf = {ratio -> 1/2}
TextString[conf]
<!-- Convert a dict-like configuration to a string -->
saveName[c_] := StringJoin[ToString /@ c]
```

>>>>

---
title: "Multifluid model of one-dimensional force-free current sheet equilibrium"
format:
  html: default
  pptx:
    slide-level: 2
    incremental: true
code-links:
  - text: Tana
    href: https://app.tana.inc?nodeid=OzzX0sVA4_6X
  - text: Zotero
    href: zotero://select/library/collections/DZHNUFNV
---

::: hide
\newcommand{\vect}[1]{\mathbf{#1}}
:::

# Introduction

Classification of discontinuities

The ratio $R_{V B}$ is expected to equal unity for rotational discontinuity, but is arbitrary for tangential discontinuities.

$$
R_{V B} = \frac{1}{\sqrt{A}} \frac{|Δ \vect{V}|}{|Δ \vect{V_A}|}
$$

where $Δ \vect{V_A} = Δ {\frac{\vect{B}}{\sqrt{\mu_0 \rho}}}$ and the anisotropy factor A is defined as $A = 1 - \frac{p_{\|}-p_{\perp}}{B^2/\mu_0}$.

## Examples

![Three examples of current sheets observed by Parker Solar Probe (PSP), ARTEMIS and Wind spacecraft with sub-Alfvenic flow velocity.](figures/fig_examples.svg)

@@

## Force-free current sheets

Force-free current sheets are important for modelling low-$β$ plasmas in laboratory, space and astrophysical applications [@marshForcefreeMagneticFields1996]

However, observations of current sheet reveals discrepancies between

- Alfven velocity and plasma velocity change across current sheet
- Anisotropic MHD theory-predicted and directly measured ion anisotropies

## Theory - previous art

### Multi-fluid model

A multi-fluid model is developed to estimate the ratio of $R_{V B}$ [@neugebauerReexaminationRotationalTangential1984]. However no analytical solution is provided for the force-free current sheet.

$$
R_{V B}(\text { multistream })=\left(\frac{1}{A} \sum_i \frac{|Δ \vect{V_i}|^2}{|Δ \vect{V_{i,A}} |^2}\right)^{1 / 2}
$$

@@

### Steinhauer, McCarthy, and Whipple (2008)

For the symmetric three-fluid system with $B_y=0$. However, it is not force-free.

@@

### Vlasov equilibrium

A number of Vlasov-Maxwell distribution function have been found to describe force-free current sheet recently

$$
\begin{aligned}
& \frac{\mathrm{d}^2 A_x}{\mathrm{d} z^2}=-\mu_0 \frac{\partial P_{z z}}{\partial A_x} \\
& \frac{\mathrm{d}^2 A_y}{\mathrm{d} z^2}=-\mu_0 \frac{\partial P_{z z}}{\partial A_y}
\end{aligned}
$$

::: {.notes}
As has been shown before, the 1D VM equilibrium equations are equivalent to the motion of a pseudoparticle in a conservative pseudopotential, with the pseudopotential being proportional to one of the diagonal components of the plasma pressure tensor.
:::

- no normal field, $B_z = 0$

# Plasma-Field Equations

## Multi-fluid collisionless plasma model in a nutshell

The governing equations for the multi-fluid collisionless plasma model can be expressed in conservation law form as follows:

$$
(\frac{∂}{∂ t} + \mathbf{u}_{α} · ∇) n_α = - n_α ∇ · \mathbf{u}_α
$$

$$
m_α n_α (\frac{∂}{∂ t} + \mathbf{u}_{α} · ∇) \mathbf{u}_α  = - ∇ · \mathbf{P}_α + q_α n_α \left( \mathbf{E} + \frac{\mathbf{u}_α \times \mathbf{B}}{c} \right)
$$

$$
\mathbf {\nabla } \times \mathbf {B} =\mu _{0}\mathbf {J}
$$ {#eq-Ampere}

$$
∇ · \mathbf{B} = 0
$$

where $α$ indicates the particle species...

## To much, let's simplify

- equilibrium solutions => $\frac{∂}{∂ t}=0$

- current sheet => magnetic field depends only on a single Cartesian coordinate, $z$

- force-free condition => $B_x^2 + B_y^2 = B_0^2 = \text{const}$

Additionally, we assume

- background electron is massless => electrons follow the magnetic field lines $\mathbf{u}_e \times \mathbf{B} = 0$

- ion populations are gyrotropic => eliminating any non-diagonal terms in the pressure tensor

To further simplify, we employ a dimensionless system by normalizing the variables with their asymptotic values:

- magnetic field by $B_{\text{ref}} = B_z$

- density by $n_{\text{ref}} = n(∞)$

## After all the simplifications

Conservation of fluid mass integrates to a constant parameter $Γ_α = n_α u_{αz}$;

Gauss's law gives another constant parameter $B_{z} = \text{const}$;

The momentum equations for each ion species are then given by:

@@

$$
d B_y / d z = - J_x = - \sum_α n_α u_{α x} + n_e u_{e x} = -n u_x + Γ_e B_x / B_z
$$ {#eq-Jx}

$$
d B_x / d z = J_y = \sum_α n_α u_{α y} - n_e u_{e y} = n u_y - Γ_e B_y / B_z
$$ {#eq-Jy}

@@

<!-- 
::: {#def-1}
Express the quantities of the second ion species relative to the first species

@@

- $Γ_e = (1 + λ_z λ_n) Λ_1 =  Λ_z  Γ_1$. 
- $Λ_z$ is constant while $Λ_x, Λ_y$ are not.
::: 
-->

@@

@@

$$
\Rightarrow 
θ_α' = - \frac{n_α B_z}{Γ_α} + \frac{B_0}{u_α}
$${#eq-theta1}

- Note that $θ_α, n_α$ are dependent variables, and $u_α, B_0, B_z, Γ_α$ are constants determined by the system.
- **So given the profile of $n_α$, we could solve the above equation to get the profile of $θ_α$, thus the profile of $u_α$.**
- The derivative of $θ$ goes to zero at infinity, gives us a relation between $u_α$ and $B_0$:

  $$
  u_α = \pm \frac{B_0 Γ_α}{B_z n_α(\infty)}
  $${#eq-u1}

@@

$$
- B_0 \sin θ θ' = \sum n_α u_{α} \sin θ_α - Γ_e B_0 \sin θ / B_z
$$

$$
\Rightarrow 
θ' = - \frac{ \sum n_α u_{α} }{B_0} + \frac{Γ_e}{B_z}
$${#eq-theta}

The above equation relates the rotation angle $θ$ to the plasma bulk velocity $n u \equiv \sum n_α u_{α}$. By equating the above two equations of $θ'$, we could get a equation relating the plasma bulk velocity to one specific species:

$$
n u = B_0 \left(\frac{Γ_e}{B_z}+\frac{n_α B_z}{Γ_α}-\frac{B_0}{u_α}\right)
$$

## Solutions

@@

Assuming density profile

$$
n_1(z) \to n_1(∞) +  \frac{κ Γ_1}{1+(z/L)^2}
$$

Solving the @eq-theta1-asym analytically, we have the rotation angle profile as

$$
\theta (z)\to \frac{\pi}{2} - \kappa  L B_z \tan ^{-1}\left(\frac{z}{L}\right)
$$

## Beutiful moment - solutions

:::: {.columns}

::: column
In dimensionless form

$$
\begin{aligned}
& B_x(z) = B_0 \sin \left(\kappa  L B_z \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
& B_y(z) = B_0 \cos \left(\kappa  L B_z \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
& n(z) = n(\infty ) + \frac{\kappa ∑_α Γ_α}{1+(z/L)^2}
\\
& J_x(z) = \frac{\kappa B_0 B_z}{1+(z/L)^2} \sin \left(\kappa  L B_z \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
& J_y(z) = \frac{\kappa B_0 B_z}{1+(z/L)^2} \cos \left(\kappa  L B_z \tan ^{-1}\left(\frac{z}{L}\right)\right)
\end{aligned}
$$
:::

::: column
In physical units

$$
\begin{aligned}
& B_x(z) = B_0 \sin \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
& B_y(z) = B_0 \cos \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
& n(z) = n_{\infty } + \frac{\kappa}{V_A} \frac{∑_α Γ_α}{1+(z/L)^2}
\\
& J_x(z) = \frac{\kappa B_0}{\mu_0 d_i}\frac{1}{1+(z/L)^2} \sin \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\\
& J_y(z) = \frac{\kappa B_0}{\mu_0 d_i}\frac{1}{1+(z/L)^2} \cos \left(\frac{\kappa L}{d_i} \tan ^{-1}\left(\frac{z}{L}\right)\right)
\end{aligned}
$$
:::
::::

@@

![](figures/J_profiles_sym.svg)

@@

@@

## Symmetric case

![](figures/profiles_sym.svg){#fig-profilesEx1}

![](figures/J_profiles_sym.svg){#fig-jprofilesEx1}

(a). Magnetic field, ion density, and ion bulk velocity for $n_1 = n_2, u_1 = -u_2$ and $δ_1=1, c_1=1, B_0 = 2$. (b). Current density profiles for the same case.

## Asymmetric case

![](figures/profiles_n1Inf=0.6.svg)

![](figures/J_profiles_n1Inf=0.6.svg)

Same as @fig-profiles, but for two ion species with same bulk velocity but different densities $n_1(∞) = 0.6$ and $B_y(z=\infty) = 1/2 B_0$.

![Plasma velocity $U_x$ profiles normalized by aysmpotic Alfvén velocity for different $n_1(∞)$.](figures/UxNormB0.svg){#fig-uxnorm}

![Plasma velocity $U_y$ profiles normalized by aysmpotic Alfvén velocity for different $n_1(∞)$.](figures/UyNormB0.svg){#fig-uynorm}

![Plasma velocity $U_x$ profiles normalized by $v_{A,x}$ for different $n_1(∞)$.](figures/UxNormBx.svg){#fig-uxNormBx}

![Plasma velocity $U_y$ profiles normalized by $v_{A,y}$ for different $n_1(∞)$.](figures/UyNormBy.svg){#fig-uyNormBy}

![](figures/vRatios.svg){#fig-vRatiosEx1}

![](figures/vRatiosEx2.svg){#fig-vRatiosEx2}

Asymptotic velocity for each species, bulk velocity and normalized one by Alfvén velocity for the case of $u_{1,z} = - u_{2,z}$ (a) and $u_{1,z} = - 2 u_{2,z}$ (b).
